---
title: "Data Wrangling"
output: html_document
date: "2024-02-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Wrangling

A task that consumes most of the analyst's time is getting the data in the right format and content. This task could include removing missing values, recoding variables, identifying noise points, choosing the right data type, etc. Once data has been coded properly, the analyst's task could involve selecting a part of the data (filtering), summarizing, and sorting by different measures. This may also include combining different datasets, creating new variables, and other data manipulation. 

R has many ways to accomplish the above. We will use [dplyr](https://dplyr.tidyverse.org/) package developed by Hadley Wickham for these tasks.    


#### Load Packages
```{r}
library(dplyr)
```

We will use the *mtcars* built in dataset to illustrate all data manipulation exercises in this section

### Subsetting or Filtering

Before we use dplyr, let us look at some basic subsetting examples

```{r}
x <- c(4, 3, 8, 2, 4) # create a vector

# Now, let's subset. Note the use of  [ ]
x[3] # This gives the third element of the vector
x[1:4] # Elements 1 through 4
x[-3] # All elements except 3
x[c(2, 5)] # The second and fifth elements

# Now, let's try conditional subsetting
x[x>3] # All values > 3
x[x> 3 & x==8] # All values > 3 and = 8. Note the use of "=="
x[x< 3 | x==4] # All values > 3 OR = 8. Note the use of "=="

# We can apply this to dataframes also

summary(mtcars)
head(mtcars)
mtcars[4:6,2] # give rows 4 through 6, for column (variable) 2
mtcars[mtcars$mpg>25,] # Choose all variables for cars with mpg > 25
mtcars[mtcars$mpg>20 & mtcars$cyl>4,1:2]
```

### Overview - dplyr

The following overview is directly from the [website](https://dplyr.tidyverse.org/)

dplyr is a grammar of data manipulation, providing a consistent set of verbs that help you solve the most common data manipulation challenges:

* mutate() adds new variables that are functions of existing variables
* select() picks variables based on their names.
* filter() picks cases based on their values.
* summarise() reduces multiple values down to a single summary.
* arrange() changes the ordering of the rows.

These all combine naturally with *group_by()* which allows you to perform any operation “by group”. You can learn more about them in *vignette("dplyr")*. As well as these single-table verbs, dplyr also provides a variety of two-table verbs, which you can learn about in *vignette("two-table")*.

If you are new to dplyr, the best place to start is the data transformation chapter in [R for Data Science](https://r4ds.hadley.nz/data-transform).

#### Subsetting

```{r}
library(dplyr)
filter(mtcars, mpg>25) # find all data for cars with mpg > 25

select(mtcars,mpg) # this selects only the mpg variable
select(mtcars,c(mpg,cyl)) # this selects mpg and cyl
select(mtcars,1:3) # selects first three variable
select(mtcars,starts_with("m")) # all variables that start with m

```

Now, let us look at some useful filterning and subsetting operations
```{r}
# remove duplicates
distinct(mtcars) 
sample_frac(mtcars, 0.5, replace = TRUE) # Sample with replacement
set.seed(4321)
sample_n(mtcars, 10, replace = TRUE) # sample 10 rows
slice(mtcars, 10:15) #Select rows by position. 
slice_max(mtcars,mpg, n=10,by=cyl)
```

Before we continue, let's take a look at an important operator, the pipe function


#### The *%>%* operator

The data pipe operator *%>%* is an important operator in producing code that is efficient and easy to read. Get comfortable in using this. This operator is part of the *magrittr* package, but is included with *dplyr*.

When you see *%>%* read it as "then" 

```{r}
mtcars$mpg %>% mean() # take mpg data and THEN calculate the mean
mtcars$mpg %>% mean() %>% sqrt() # calculate the square root of the mean mpg
myResults <- mtcars$mpg %>% mean() %>% sqrt() # stores the results
myResults
```

We can combine the filter and select operations through the pipe function
```{r}
mtcars %>%  select(cyl) %>% filter(mpg > 25) # show the cyl of all cars with mpg > 25
```



Let us now look at an extensive example to demonstrate the superiority of the pipe operator
```{r}
# What does this code do?

mtcars %>%
  filter(carb>1) %>%
  group_by(cyl) %>%
  summarise(Avg_mpg = mean(mpg)) %>%
  arrange(desc(Avg_mpg))
```

#### Grouping Data by Categorical Variables

Often, observations are nested within groups or variables, and our goal is perform statistical analysis both at the observation level and the group level. The *group_by()* function allows us to do that.
```{r}
group.cyl <- mtcars %>% group_by(cyl)
group.cyl
group.cyl %>% summarise(mean(mpg))

# Let's combine multiple steps
mtcars %>% 
  group_by(cyl) %>% 
  summarise(mean(mpg),mean(hp))
```
#### Arranging Variables by Value

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarise(Mean_mpg = mean(mpg, na.rm = T),
            Mean_hp = mean(hp, na.rm = T)) %>% 
  arrange(desc(Mean_mpg))

```

#### Grouping and Summary Statistics
Grouping data and performing summary statistics is an important part of any data analysis, so get
comfortable doing this
```{r}
mtcars %>% summarise(Mean_mpg = mean(mpg, na.rm=T))

mtcars %>% 
  group_by(cyl) %>% 
  summarise(Min = min(mpg, na.rm = T),
            Max = max(mpg, na.rm = T),
            Mean = mean(mpg, na.rm = T),
            SD = sd(mpg, na.rm = T)) %>% 
  arrange(Min)

# What if we only wanted 6 or 4 cylinder cars
mtcars %>% 
  filter(cyl<8) %>% 
  group_by(cyl) %>% 
  summarise(Min = min(mpg, na.rm = T),
            Max = max(mpg, na.rm = T),
            Mean = mean(mpg, na.rm = T),
            SD = sd(mpg, na.rm = T)) %>% 
  arrange(Min)
```

#### Creating New Variables

We will use the mutate function to create new variables
```{r}
mpg_cyl <- mtcars %>% mutate(mc = mpg / cyl)
head(mpg_cyl)
```

#### Using sapply
sapply( ) is a function can be used to apply a given function to all columns of
a data frame in an efficient and elegant fashion.
```{r}
mtcars %>%  select(mpg,disp,hp) %>% sapply(sd)
```


### Help

* [Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
