---
title: "Making Graphs"
author: "Murali Shanker"
date: "9/20/2021"
output:
  html_document: default
  toc: yes
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Graphs

In R, we can *build* graphs one statement at a time

```{r}
attach(mtcars)
plot(wt, mpg)
abline(lm(mpg~wt))
title("Regression of MPG on Weight")
detach(mtcars)
```

## An Example
The following table describes patient to two drugs:

Dosage | Drug A | Drug B
------- | ------ | ---------
20 | 16 | 15
30 | 20 | 18
40 | 27 | 25
45 | 40 | 31
60 | 60 | 40

```{r}
dose  <- c(20, 30, 40, 45, 60)
drugA <- c(16, 20, 27, 40, 60)
drugB <- c(15, 18, 25, 31, 40)

drug <- data.frame(dose,drugA,drugB)

plot(dose, drugA, type="b") # type b indicates points and lines

```
### Parameters

Many features of the graph can be modified using the *par()* function. The format is *par(optionname=value, optionname=value, ...)*. Specifying par() without parameters produces a list of the current graphical settings. Adding the no.readonly=TRUE option produces a list of current graphical settings that can be modified.

```{r}
opar <- par(no.readonly=TRUE) # make a copy of current settings
par(lty=2, pch=17) # change line type to dashed, and points to solid triangle
plot(dose, drugA, type="b")
par(opar) # restore original setting

# Alternatively,
plot(dose, drugA, type="b", lty=2, pch=17)
```

The following table specifies the parameters for symbols and lines

Parameter | Description
--------- | ---------------
pch | Specifies the symbol to use when plotting points
cex | Specifies the symbol size. cex is a number indicating the amount by which plotting symbols should be scaled relative to the default. 1 = default, 1.5 is 50% larger, 0.5 is 50% smaller, and so forth.
lty | Specifies the line type 
lwd | Specifies the line width. lwd is expressed relative to the default (1 = default). For example, lwd=2 generates a line twice as wide as the default.

## Easier Option

You will normally plot graphs using ggplot. To help you understand ggplot better, and also provide an easier way to do plots, we will use a package.

```{r}
# installed.packages("ggplot2")
library("ggplot2")

#install.packages("esquisse")
library("esquisse")
#esquisser(drug)

```

 
