---
title: "Linear Regression"
output: html_document
date: "2024-04-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simple Linear Regression

```{r}
library(caret)
library(dplyr)
```

### Example
```{r}
X=c(61,63,67,69,70,74,76,81,86,91,95,97)
Y=c(4.28,4.08,4.42,4.17,4.48,4.3,4.82,4.7,5.11,5.13,5.64,5.56)
```

We use the `lm` command to run linear regression
```{r}
Model = lm(Y~X)
summary(Model)
plot(Model$fitted.values,Model$residuals)
plot(X,Y)
```
### Another Example
```{r}
library(ISLR)
Model = lm(Sales~Price,data=Carseats)
plot(Model$fitted.values,Model$residuals,xlab="fitted values",ylab="Residuals")
hist(Model$residuals)
qqnorm(Model$residuals,col="red")
qqline(Model$residuals)
```

## Multiple Linear Regression

### Problem Description
The city tax assessor was interested in predicting residential home sales prices in a midwestern city as a function of various characteristics of the home and surrounding property. 

Data on 522 arms-length transactions were obtained for home sales during the year 2002. 

Each line of the data set has an identification number and provides information on 12 other 
variables. The 13 variables are: 


Variable Number | Variable Name | Description
----------------| ------------- | ------------
1 | Identification number | 1-522
2 | Sales price | Sales price of residence (dollars) 
3 | Finished square feet | Finished area of residence (square feet) 
4 | Number of bedrooms | Total number of bedrooms in residence
5 | Number of bathrooms | Total number of bathrooms in residence
6 | Air conditioning | Presence or absence of air conditioning: 1 if yes; 0 otherwise 
7 | Garage size | Number of cars that garage will hold
8 | Pool | Presence or absence of swimming pool: 1 if yes; 0 otherwise 
9 | Year built 
10 | Quality | Index for quality of construction: 
1 indicates high quality; 
2 indicates medium quality; 
3 indicates low quality 
11 | Style | Qualitative indicator of architectural style
12 | Lot size | in Square Feet
13 | Adjacent to highway | Presence or absence of adjacency to highway: 1 if yes; 0 otherwise

#### Data
```{r}
realestate <- read.csv("C:/Users/mshan/Downloads/APPENC07.csv")
head(realestate)
```
Now, let's construct some models
```{r}
Model = lm(SalesPrice ~SqFt+Bedrooms+Bathrooms,realestate)
summary(Model)

```

